Using functor categories to generate intermediate code with Agda

### 2025-04-23 draft.tex
programa example added; citation adjusted

\definecolor{mediumblue}{HTML}{0000CD}
\definecolor{green}{HTML}{008B00}
\newcommand{\mb}[1]{\textcolor{mediumblue}{#1}}
\newcommand{\gn}[1]{\textcolor{green}{#1}}

@@

    This dissertation presents an implementation of a compiler from the source language to the target language with Agda \cite{Agda}. This implementation is based on the work of Reynolds \cite{Reynolds}, who presented a denotational semantics of Algol-like languages in the form of a presheaf category over stack descriptors. The compiler is implemented as a functor from the source language to the target language. The implementation is verified with Agda's type system, which ensures that the generated code is well-typed and adheres to the semantics of both the source and target languages. This implementation proves and refines Reynolds' work, providing a practical example of how to use functor categories to generate intermediate code.

    \section{Motivation and related work} \label{sec: motivation}
        The denotational semantics of Algol-like languages can be structured as a presheaf category over stack descriptors, which has been shown by Reynolds \cite{essence} and Oles \cite{Oles_1} \cite{Oles_2}. By interpreting the source language into this category, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler. The mathematical structure of the compiler has been specified by Reynolds in his paper ``Using Functor Categories to Generate Intermediate Code'' \cite{Reynolds}.

@@

        \motivation{Trend of verified compilers}
        The rise of verified compilers including CompCert \cite{CompCert}, CakeML \cite{CakeML} reflects a broader trend toward trustworthy systems, where correctness proofs replace testing for critical guarantees. Like Lean 4 \cite{Lean4}, we leverage dependent types to internalise the verification of correctness of terms.

    \section{Language choice: Agda's advantages}
        Agda \cite{Agda} is a dependently typed programming language and proof assistant. Agda captures the source language's intrinsic sytax with indexed families, which contains only well-typed terms. Therefore, it focuses on the correct programs and rules out the ill-typed nonsensical inputs.

        Dependently typed languages provide a natural framework for expressing functor categories is proven both theoretically and practically. There have been dependent-type-theoretic model of categories \cite{Dybjer}, and it has been shown that functor categories arise naturally as dependent function types \cite{Jacobs}. A formalisation of Category Theory, including cartesian closed categories, functors and presheaves has been developed in Agda by Hu and Caratte \cite{Cat_Agda}. Other proof assistants, such as Isabelle/Hol, does not have a dependently typed language structure, and thus cannot express the functor categories as naturally as Agda. Agda also provides an interactive environment for writing and verifying programs, which will be further discussed in \secref{subsec: holes}.

@@

    \section{Starting Point}
    Prior to this project, I had no experience with Agda. Although I was aware of the open-source online tutorial \textit{Programming Language Foundations in Agda} (PLFA) \cite{plfa}, my preparation was limited to setting up the Agda environment on my laptop by following the ``Front Matter'' section of the tutorial.

@@

    \section{Category theory background} \label{sec: cat}
        Category theory provides a high-level abstraction from which we can reason about the structure of mathematical objects and their relationships. It provides us a ``bird's eye view'' of the mathematics which enable us to spot patterns that are difficult to see in the details. \cite{basic_cat} More specifically, it provides a ``purer'' view of functions that is not derived from sets. \cite{scott-lambda} Compared to set theory which is ``element-oriented'', category theory is ``function-oriented'' and ``morphism-oriented''. We understand structures not via elements but by how they transform into each other.

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory. Curry-Howard correspondence \cite{curry-howard} states that there is a correspondence between logic and type theory, where propositions correspond to types and proofs correspond to terms. This correspondence was later extended by Joachim Lambek, who showed that cartesian closed categories provide a natural semantic setting for the simply typed lambda calculus. \cite{lambek} Under this perspective, types become objects in a category, and programs correspond to morphisms. We will further explore this connection in \secref{subsec: curry-howard-lambek}.

        The following is a brief introduction to the basic concepts of category theory, which is based on the work of Leinster \cite{basic_cat}, 
        Riehl \cite{cat_context}, and the lecture notes of Part II Category Theory by Andrew Pitts and Marcelo Fiore \cite{cat_lecture_notes}.

@@

    \section{Agda}
        \subsection{Basic datatypes and pattern matching}
        We will go through a simple example in PLFA \cite{plfa} to illustrate the basic datatypes and pattern matching in Agda.

        \[\includegraphics[width=\textwidth]{example.pdf}\]

@@

        \subsection{Standard library}
        The Agda standard library \cite{agda_std} is a collection of modules that provide a wide range of useful functions and types. It includes modules for basic data types, such as natural numbers, lists, and vectors. In my implementation, however, I defined most of the basic data types and functions from scratch for the following two reasons:
        \begin{itemize}
            \item
                I wanted to have a better understanding of the basic data types and functions, so I decided to implement them from scratch.
            \item
                I need a particular representation of minus operation that ensures the input is always valid. This is further explained in [to add]
        \end{itemize}

        With standard library, we can use actual numbers $1, 2, 3, \dots$ instead of calling the constructor $\gn{\mathsf{suc}}$ multiple times. This is a convenient feature for testing. Agda provides a way to link self-defined functions to the standard library with a BUILTIN programa. This allows me to use a convenient syntax for numbers, while still using the self-defined functions in the implementation.

@@

    \section{Tools Used}
    Completing the project is an iterative process. I used Git \cite{git} for version control, and work had been synchronised with a GitHub \cite{github} repository for backup.

    For the development environment, I tried both Emacs \cite{emacs} and Visual Studio Code \cite{vscode} with an agda-mode extension \cite{agda_mode} on Windows Subsystem for Linux with Ubuntu \cite{wsl_ubuntu} 22.04 LTS. I am more familiar with the snippet and syntax highlighting features of Visual Studio Code, so I used it for most of the development.

    Code from the PLFA tutorial and Agda standard library \cite{agda_std} were used as references.

>>>>

\definecolor{mediumblue}{HTML}{0000CD}
\definecolor{green}{HTML}{008B00}
\definecolor{orange}{HTML}{CD6600}

\newcommand{\mb}[1]{\textcolor{mediumblue}{#1}}
\newcommand{\gn}[1]{\textcolor{green}{#1}}
\newcommand{\og}[1]{\textcolor{orange}{#1}}

@@

    This dissertation presents an implementation of a compiler from the source language to the target language with Agda~\cite{Agda}. This implementation is based on the work of Reynolds~\cite{Reynolds}, who presented a denotational semantics of Algol-like languages in the form of a presheaf category over stack descriptors. The compiler is implemented as a functor from the source language to the target language. The implementation is verified with Agda's type system, which ensures that the generated code is well-typed and adheres to the semantics of both the source and target languages. This implementation proves and refines Reynolds' work, providing a practical example of how to use functor categories to generate intermediate code.

    \section{Motivation and related work} \label{sec: motivation}
        The denotational semantics of Algol-like languages can be structured as a presheaf category over stack descriptors, which has been shown by Reynolds~\cite{essence} and Oles~\cite{Oles_1}~\cite{Oles_2}. By interpreting the source language into this category, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler. The mathematical structure of the compiler has been specified by Reynolds in his paper ``Using Functor Categories to Generate Intermediate Code''~\cite{Reynolds}.

@@

        \motivation{Trend of verified compilers}
        The rise of verified compilers including CompCert~\cite{CompCert}, CakeML~\cite{CakeML} reflects a broader trend toward trustworthy systems, where correctness proofs replace testing for critical guarantees. Like Lean 4~\cite{Lean4}, we leverage dependent types to internalise the verification of correctness of terms.

    \section{Language choice: Agda's advantages}
        Agda~\cite{Agda} is a dependently typed programming language and proof assistant. Agda captures the source language's intrinsic sytax with indexed families, which contains only well-typed terms. Therefore, it focuses on the correct programs and rules out the ill-typed nonsensical inputs.

        Dependently typed languages provide a natural framework for expressing functor categories is proven both theoretically and practically. There have been dependent-type-theoretic model of categories~\cite{Dybjer}, and it has been shown that functor categories arise naturally as dependent function types~\cite{Jacobs}. A formalisation of Category Theory, including cartesian closed categories, functors and presheaves has been developed in Agda by Hu and Caratte~\cite{Cat_Agda}. Other proof assistants, such as Isabelle/Hol, does not have a dependently typed language structure, and thus cannot express the functor categories as naturally as Agda. Agda also provides an interactive environment for writing and verifying programs, which will be further discussed in \secref{subsec: holes}.

@@

    \section{Starting Point}
    Prior to this project, I had no experience with Agda. Although I was aware of the open-source online tutorial \textit{Programming Language Foundations in Agda} (PLFA)~\cite{plfa}, my preparation was limited to setting up the Agda environment on my laptop by following the ``Front Matter'' section of the tutorial.

@@

    \section{Category theory background} \label{sec: cat}
        Category theory provides a high-level abstraction from which we can reason about the structure of mathematical objects and their relationships. It provides us a ``bird's eye view'' of the mathematics which enable us to spot patterns that are difficult to see in the details~\cite{basic_cat}. More specifically, it provides a ``purer'' view of functions that is not derived from sets~\cite{scott-lambda}. Compared to set theory which is ``element-oriented'', category theory is ``function-oriented'' and ``morphism-oriented''. We understand structures not via elements but by how they transform into each other.

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory. Curry-Howard correspondence~\cite{curry-howard} states that there is a correspondence between logic and type theory, where propositions correspond to types and proofs correspond to terms. This correspondence was later extended by Joachim Lambek, who showed that cartesian closed categories provide a natural semantic setting for the simply typed lambda calculus~\cite{lambek}. Under this perspective, types become objects in a category, and programs correspond to morphisms. We will further explore this connection in \secref{subsec: curry-howard-lambek}.

        The following is a brief introduction to the basic concepts of category theory, which is based on the work of Leinster~\cite{basic_cat}, 
        Riehl~\cite{cat_context}, and the lecture notes of Part II Category Theory by Andrew Pitts and Marcelo Fiore~\cite{cat_lecture_notes}.

@@

    \section{Agda}
        \subsection{Basic datatypes and pattern matching}
        We will go through a simple example in PLFA~\cite{plfa} to illustrate the basic datatypes and pattern matching in Agda.

        \[\includegraphics[width=\textwidth]{natural_plus.pdf}\]

@@

        \subsection{Standard library}
        The Agda standard library~\cite{agda_std} is a collection of modules that provide a wide range of useful functions and types. It includes modules for basic data types, such as natural numbers, lists, and vectors. In my implementation, however, I defined most of the basic data types and functions from scratch for the following two reasons:
        \begin{itemize}
            \item
                I wanted to have a better understanding of the basic data types and functions, so I decided to implement them from scratch.
            \item
                I need a particular representation of minus operation that ensures the input is always valid. This is further explained in [to add]
        \end{itemize}

        With standard library, we can use actual numbers $1, 2, 3, \dots$ instead of calling the constructor $\gn{\mathsf{suc}}$ multiple times. This is a convenient feature for testing. Agda provides a way to link self-defined functions to the standard library with a \og{\textsf{BUILTIN}} programa. This allows me to use a convenient syntax for numbers, while still using the self-defined functions in the implementation.
        \[\includegraphics[width=\textwidth]{natural_programma.pdf}\]
        Note that in agda we can use unicode characters for terms, which makes the code more readable. For example, here we simply name the term \mb{1+1$\equiv$2}.

@@

    \section{Tools Used}
    Completing the project is an iterative process. I used Git~\cite{git} for version control, and work had been synchronised with a GitHub~\cite{github} repository for backup.

    For the development environment, I tried both Emacs~\cite{emacs} and Visual Studio Code~\cite{vscode} with an agda-mode extension~\cite{agda_mode} on Windows Subsystem for Linux with Ubuntu~\cite{wsl_ubuntu} 22.04 LTS. I am more familiar with the snippet and syntax highlighting features of Visual Studio Code, so I used it for most of the development.

    Code from the PLFA tutorial and Agda standard library~\cite{agda_std} were used as references.
